Follow-up Samples

### 2017-01-16 submission/manuscript_outline_20161024.Rmd
streamline discussion

Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.  This data suggests that what may change after surgery is not necessarily any one specific bacterium nor any addition of new or depressed bacterium but rather that  a shift of the bacterial community as a whole occurs with the existing members of this microbiome.

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial and follow up samples in order to confirm that the community was what was responding or chaning due to surgery.  We found that the OTUs that made up most important variables to the model overwhelmingly belonged to comensal bacteria.  With only the lesion model having a single OTU from previously associated cancer bacteria (Porphyromonas asaccharolytica).  Using only these important OTUs and FIT both models (lesion and initial and follow up) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Confirming the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and all belonged to regular residents of our gut community.

There was no difference for the majority of models tested for differences in positive probablity based on whether chemotherapy or radiation was received.  There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.  These results would indicate that the findings described on specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to see E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  One potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  An alternative reason would be that even if they were reduced in relative abundance they were still present.  Another potential explanation is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Finally, it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  It is also possible that all three of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery.  It should be stated that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or initial and follow up).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the fecal DNA test [@cotter_long-term_2016] could help improve our overall AUC.  Another limitation is that this study drew heavily from those with caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

Within figure 3 the initial and follow up model showed better test AUC results then the training set AUC.  This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be overfit since the total number of samples is not that large.  However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with fecal, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

>>>>

Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.  This data suggests that the changes that may be important in the bacterial microbiome after surgery are not necessarily any one specific bacterium, a clear depression of CRC related bacteria, nor any addition of new or depressed bacterium but rather that a shift of the bacterial community as a whole occurs with the existing members of this microbiome.

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial and follow up samples in order to confirm that the community was what was responding or changing due to surgery.  We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.  With only the lesion model having a single OTU from previously associated cancer bacteria (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and initial and follow up) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Confirming the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

There was no difference for the majority of models tested for differences in positive probablity based on whether chemotherapy or radiation was received.  There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.  These results would indicate that the findings described were specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  One potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  An alternative reason would be that even if they were reduced in relative abundance they were still present.  Another potential explanation is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  It is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all three of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be stated that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or initial and follow up).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the fecal DNA test [@cotter_long-term_2016] could help improve our overall AUC.  Another limitation is that this study drew heavily from those with caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

Another interesting outcome was that Within figure 3 the initial and follow up model showed better test AUC results then the training set AUC.  This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be overfit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

### 2017-01-16 submission/manuscript_outline_20161024.tex
streamline discussion

Interestingly, when only previously associated cancer bacteria were
investigated only 2 out of the 4 had a significant decrease in relative
abundance between initial and follow up for the carcinoma group and 0
out of the 4 were significant for the adenoma group. This data suggests
that what may change after surgery is not necessarily any one specific
bacterium nor any addition of new or depressed bacterium but rather that
a shift of the bacterial community as a whole occurs with the existing
members of this microbiome.

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or initial and follow up samples in order to confirm that the community
was what was responding or chaning due to surgery. We found that the
OTUs that made up most important variables to the model overwhelmingly
belonged to comensal bacteria. With only the lesion model having a
single OTU from previously associated cancer bacteria (Porphyromonas
asaccharolytica). Using only these important OTUs and FIT both models
(lesion and initial and follow up) had a significantly decreased
positive probability of either lesion or being an initial sample on
follow up {[}Figure 4 \& S4{]}. Confirming the importance of the changes
of comensal bacteria to these classifications was that a total of 14
OTUs were common to both models and all belonged to regular residents of
our gut community.

There was no difference for the majority of models tested for
differences in positive probablity based on whether chemotherapy or
radiation was received. There was also no difference in the thetayc
distance metric based on length of time between initial and follow up
sample between adenoma and carcinoma. These results would indicate that
the findings described on specific to the surgical intervention and that
differences observed between carcinoma and adenoma samples can not be
simply attributed to collection time between samples.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool {[}13,14{]}
by exploring what happens to the bacterial community after surgical
removal of a lesion. Based on previous work by Arthur, et al. {[}18{]}
it may not be surprising to see E.coli as one of the most important OTUs
and one that was common to both models. Interestingly, many of the most
important OTUs had taxonomic identification for resident gut microbes.
This could suggest that the bacterial community is one of the first
components that could change during the pathogenesis of disease. These
bacterial microbiome changes could be the first step in allowing more
inflammatory bacterium to gain a foothold within the colon {[}19{]}.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. One potential explanation is that even
with surgery and a shift of the bacterial community these specific
bacteria still persisted within the colon. An alternative reason would
be that even if they were reduced in relative abundance they were still
present. Another potential explanation is that they were not present in
enough individuals to be able to classify those with and without disease
with a high degree of accuracy. Finally, it is possible that our Random
Forest models were able to get the same information from measures such
as FIT or other OTUs. It is also possible that all three of these
potential explanations could have played a role. Regardless, our
observations would suggest that an individual's resident bacteria have a
role to play in disease initiation and could change in a way that allows
predictive models to lower the positive probability of a lesion after
surgery. It should be stated that our study does not argue against the
importance of these CRC associated bacteria in the pathogenesis of
disease but rather that the models do not utilize these specific
bacteria for classification purposes (lesion or initial and follow up).
In fact, it is possible that these CRC associated bacteria are important
in the transition from adenoma to carcinoma and would be one explanation
as to why in our data we not only see high initial relative abundances,
in certain individuals, but also large decreases in relative abundance
in those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our Random Forest based model keeping a number of
individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having a lesion. Another limitation is that
we do not know if adding modern tests such as the fecal DNA test
{[}20{]} could help improve our overall AUC. Another limitation is that
this study drew heavily from those with caucasian ancestry. The results
may not be immediately representative of those with either Asian or
African ancestry. Finally, although our training and test set are
relatively large we still run the risk of overfitting or having a model
that may not be immediately extrapolateable to other populations. We've
done our best to safeguard against this by not only running 10-fold
cross validation but also having over 100 different 80/20 splits to try
and mimic the type of variation that might be expected to occur.

Within figure 3 the initial and follow up model showed better test AUC
results then the training set AUC. This may have occured because the
training AUC that was determined from 20 repeated 10 fold cross
validation removed samples at random and did not take into account that
they were matched samples. Another potential reason is that the model
itself may be overfit since the total number of samples is not that
large. However, the lesion model did not suffer from these discrepencies
and similar conclusions can be drawn solely from this model. Regardless,
further independent studies will need to be carried out to verify our
findings since not only are we dealing with fecal, which could be very
different than the communities present on the actual tissue, but also
are dealing with correlations that may not be representative of the true
pathogensis of disease.

>>>>

Interestingly, when only previously associated cancer bacteria were
investigated only 2 out of the 4 had a significant decrease in relative
abundance between initial and follow up for the carcinoma group and 0
out of the 4 were significant for the adenoma group. This data suggests
that the changes that may be important in the bacterial microbiome after
surgery are not necessarily any one specific bacterium, a clear
depression of CRC related bacteria, nor any addition of new or depressed
bacterium but rather that a shift of the bacterial community as a whole
occurs with the existing members of this microbiome.

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or initial and follow up samples in order to confirm that the community
was what was responding or changing due to surgery. We found that the
OTUs that made up the most important variables to the model
overwhelmingly belonged to comensal bacteria. With only the lesion model
having a single OTU from previously associated cancer bacteria
(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
and FIT both models (lesion and initial and follow up) had a
significantly decreased positive probability of either lesion or being
an initial sample on follow up {[}Figure 4 \& S4{]}. Confirming the
importance of the changes of comensal bacteria to these classifications
was that a total of 14 OTUs were common to both models and the vast
majority belonged to regular residents of our gut community.

There was no difference for the majority of models tested for
differences in positive probablity based on whether chemotherapy or
radiation was received. There was also no difference in the thetayc
distance metric based on length of time between initial and follow up
sample between adenoma and carcinoma. These results would indicate that
the findings described were specific to the surgical intervention and
that differences observed between carcinoma and adenoma samples can not
be simply attributed to collection time between samples.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool {[}13,14{]}
by exploring what happens to the bacterial community after surgical
removal of a lesion. Based on previous work by Arthur, et al. {[}18{]}
it may not be surprising to have E.coli as one of the most important
OTUs and one that was common to both models. Interestingly, many of the
most important OTUs had taxonomic identification for resident gut
microbes. This could suggest that the bacterial community is one of the
first components that could change during the pathogenesis of disease.
These bacterial microbiome changes could be the first step in allowing
more inflammatory bacterium to gain a foothold within the colon
{[}19{]}.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. One potential explanation is that even
with surgery and a shift of the bacterial community these specific
bacteria still persisted within the colon. An alternative reason would
be that even if they were reduced in relative abundance they were still
present. Another potential explanation is that they were not present in
enough individuals to be able to classify those with and without disease
with a high degree of accuracy. It is possible that our Random Forest
models were able to get the same information from measures such as FIT
or other OTUs. Finally, it is also possible that all three of these
potential explanations could have played a role. Regardless, our
observations would suggest that an individual's resident bacteria have a
large role to play in disease initiation and could change in a way that
allows predictive models to lower the positive probability of a lesion
after surgery {[}Figure 4{]}. It should be stated that our study does
not argue against the importance of these CRC associated bacteria in the
pathogenesis of disease but rather that the models do not utilize these
specific bacteria for classification purposes (lesion or initial and
follow up). In fact, it is possible that these CRC associated bacteria
are important in the transition from adenoma to carcinoma and would be
one explanation as to why in our data we not only see high initial
relative abundances, in certain individuals, but also large decreases in
relative abundance in those with carcinoma but not in those with adenoma
after surgery {[}Figure 2{]}.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our Random Forest based model to have kept a number of
individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having a lesion. Another limitation is that
we do not know if adding modern tests such as the fecal DNA test
{[}20{]} could help improve our overall AUC. Another limitation is that
this study drew heavily from those with caucasian ancestry. The results
may not be immediately representative of those with either Asian or
African ancestry. Finally, although our training and test set are
relatively large we still run the risk of overfitting or having a model
that may not be immediately extrapolateable to other populations. We've
done our best to safeguard against this by not only running 10-fold
cross validation but also having over 100 different 80/20 splits to try
and mimic the type of variation that might be expected to occur.

Another interesting outcome was that Within figure 3 the initial and
follow up model showed better test AUC results then the training set
AUC. This may have occured because the training AUC that was determined
from 20 repeated 10 fold cross validation removed samples at random and
did not take into account that they were matched samples. Another
potential reason is that the model itself may be overfit since the total
number of samples was not that large. However, the lesion model did not
suffer from these discrepencies and similar conclusions can be drawn
solely from this model. Regardless, further independent studies will
need to be carried out to verify our findings since not only are we
dealing with feces, which could be very different than the communities
present on the actual tissue, but also are dealing with correlations
that may not be representative of the true pathogensis of disease.
